Gsmodutils: A python based framework for test driven genome scale metabolic model development

### 2018-08-13 article/article.tex
Article authors (added Charlie)
Article draft changes

\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\usepackage{color}
\usepackage{authblk}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{hyperref}
%PREPRINT/MANUSCRIPT OPTIONS
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\doublespacing

@@

\title{Gsmodutils: A python based framework for test driven genome scale metabolic model development}
\author[1]{James P Gilbert}
\author[1]{Nicole Pearcy}
\author[1]{Rupert Norman}
\author[2]{Jamie Twycross}
\author[3]{John King}
\author[1]{Nigel Minton}
\affil[1]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\affil[2]{Intelligent Modelling and Analysis group, School of Computer Science, University of Nottingham, Nottingham, NG8 1BB, United Kingdom}
\affil[3]{School of Mathematical Sciences, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}

@@

\section{Introduction}
Stoichiometric constraints based modelling for biological systems has been a mainstay of systems biology for several decades \cite{fell1986fat, varma1994stoichiometric}.
Given its flexibility, low barrier to entry and requirement only on minimal knowledge regarding the stoichiometry of metabolic networks this structural approach has become an extremely popular method for modelling steady state behaviour of large, biochemical networks \cite{kauffman2003advances}.
Such large scale reconstructions are often refereed to as genome scale metabolic models (GSMMs), as the processes is significantly aided through the advent of relatively inexpensive genome sequencing \cite{o2015using, land2015insights}.
Indeed, owing to their ability to model complex aspects of metabolism, GSMMs have been widely adopted as a standard to elucidate and optimise of industrial biotechnology processes \cite{kim2017current}.

@@

Furthermore, as with many areas of bioinformatic study the proliferation of available computational tools has become somewhat vast.
This covers huge variety of software platforms including the COBRA toolbox in Matlab \cite{schellenberger2011quantitative}, ScrumPy and COBRApy in Python \cite{poolman2006scrumpy, ebrahim2013cobrapy} with additional tools and libraries such as cameo \cite{cardoso2017cameo}, OptFlux in Java \cite{rocha2010optflux}, SurreyFBA \cite{gevorgyan2010surreyfba}.
Whilst most of these tools are Open Source and follow standards, such as SBML \cite{finney2003systems}, there are often few options that make analysis easy to replicate.

The archetype design, build and test cycle of synthetic biology heavily relies on the usage of bioinformatics software and modelling to improve the production of natural products \cite{carbonell2016bioinformatics}.
In order to speed up the use of bioinformatics tools to produce high value platform chemicals, genome scale models are often used to discover methods for process optimisation \textit{in silico}.

For example, many tools such as RetroPath \cite{carbonell2013retropath}, XTMS \cite{carbonell2014xtms} and GEM-Path \cite{campodonico2014generation} suggest thousands of potential hetrologous pathways as well as providing predictions for flux distributions.
Many of these tools significantly increase the value of genome scale models, for example by coupling commodity production to an organism's growth \cite{feist2010model}.
These tools all suggest major changes to wild-type strains must be tracked and compared to allow models to remain relevant.
In effect, mechanisms are required to related modified models to modified test and production strains.

@@

The design phase of industrial synthetic biotechnology, due to historical and regulatory convention, often follows a waterfall model for development of genome scale models.
This is borne out of a number significant constraints in the development of genome scale models.
The waterfall approach for model is characterised in Figure \ref{fig:methodology_approaches} where requirements for a model are captured at an early stage and are hard to refine during the production process.

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
Constraints based models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes often being left to a few of critical interest \cite{seemann2014prokka}.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady state phenomena still provides a good understanding of systemic properties \cite{o2015using}.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{Waterfall.png}
\caption{Conventional waterfall methodology for model development.
Each process is considered to be an isolated aspect of model development.}
\label{fig:methodology_approaches}
\end{figure}

@@

We feel that while many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions.
However, their is a major gap in terms of the reliability and re-usability of models due to a lack of standardisation and software tools to aid such processes.
The following sections provide an overview of the gsmodutils software framework.
Gsmodutils aims to provide a basis for test driven, version controlled agile model development.
All software is open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

By automatically integrating the COBRApy \cite{ebrahim2013cobrapy} users can easily write convenient test cases following examples given in the user guide.
A standard test case, ensuring that a given model grows on media is given in Box \ref{fig:test_case}.
When a new model repository is created with the gsmodutils tool, a number of pre-written test cases are automatically added to a file.
However, we stress that the vast majority of individual use cases for a model must be specific to a given biological problem.

The software provides a number of features such as import an export of models in different formats and the generation of test reports through use of the command line. 
The use of flat files for easy integration with version control software such as git and mercurial.

In addition, projects are easy to export using portable standardised docker images \cite{merkel2014docker}. 
The idea being to allow users to share models as quickly and easily as possible without concern for custom system configurations.

@@

\subsection{Strain designs}
A core aspect behind the implementation of gsmodutils is the concept of a \textit{design}, this encompasses a simple set of changes to a ``wild-type'' model that are required for analysis.
For example a mutant strain that has undergone several gene deletions is not sufficiently different from the original wild-type to merit having its own genome scale model.
However, it is often the case that such deletions are of scientific or industrial interest and, as such, the strain will be used in future work.
Consequently, such designs are hereditary in nature.
By taking the delta between the constraints applied to an initial model and subsequent modifications, gsmodutils allows users to easily reuse and export models with this design.
As designs inherit from a base model, future curation to a wild type model will automatically be included in the designs.
Similarly, designs are self contained and will not interfere with one another allowing a form of project management.
Figure \ref{fig:strain_hered} shows how this could work in a practical situation.
Designs of this nature can also be programmatic, allowing the implementation of features such as non-standard constraints that can be dynamically loaded.
Furtheremore, integrity of all strain designs is automatically tested as part of the default gsmodutils testing framework.

\begin{figure}
  \includegraphics[width=\textwidth]{inheritence.png}
  \label{fig:strain_hered}
  \caption{Examples of gsmodutils design inheritance.
  Each design stores the delta between the wild-type base model, any parents and the changes to constraints the design contains.
  In the example presented above a hetrologous production pathway is combined with a reusable set of knock-outs.
  In practice, ideally, these designs should relate to real constructs and strains evaluated by wet lab biologists.
  }
\end{figure}

\subsection{Development workflow}
In this section we propose a method for the development of genome scale models that integrates gsmodutils with version control systems.
Figure \ref{fig:test_case} highlights the notion of test cases, taken from test driven development.
The outlined workflow is that the user writes a formal test case for some modelling goal, perhaps driven by captured experimental data, that fits a specific form of validation criteria.
We note that, in principle, this test case should be written \textit{a priori} of changes to a model.

% \section{Example usage Clostridium acetobutylicum models}
% \subsection{Scientific background}
% \textit{C. acetobutylicum} ATCC 824 is a widely researched organism that has found use in the production of solvents for decades.
% More recently, \textit{C. acetobutylicum} has been applied for the production of alcohols such as ethanol and n-butanols.
% The industrial relevance and strong background of experimental data has led to the application of many genome scale models \cite{}.
% Whilst increasingly informative when combined with experimental data, the history of the management of models for \textit{C. acetobutylicum} tells a different story.
% 
% Initially, two rival models were published containing a similar number of genes \cite{lee2008genome, senger2008genome}.
% % \textbf{TODO: describe in more detail}.
% %Discuss limitations of rival models
% This led to the further development of subsequent models, iCaC 490 \cite{mcanulty2012genome} and an automatically generated model from  the Model SEED  pipeline \cite{}.
% 
% In this example we focus on the more recent contributions of iCac802 \cite{dash2014capturing} and iCac967 \cite{yoo2015quantitative}.
% iCac802 combines a significantly updated set of annotations from previously curated model versions, an improved TCA cycle, full proton balancing in reactions and reactions for more complete coverage of hexadecanoyl-acp, hexadecanoyl-CoA, crotonyl-CoA pathways as well as pathways for the uptake and biosynthesis of more compounds.
% Crucially, the model itself received significant validation against fluxonomic data from 13C-metabolic flux analysis ($13$C-MFA).
% This additional validation set considerable flux bounds upon pathways, leading to improved curation.
% 
% Whilst iCac802 was informed by previous models, critically, it was started as a new curation project, a significant undertaking of work.
% Such an approach may be necessary to uncover errors or oversights in construction processes.
% However, this naturally adds a significant time burden for model curators and may be due to a lack of adaptability from previous models.
% 
% The iCac 967 model presents a smaller set of changes to the annotation of the \textit{C. acetobutylicum} model.
% While the number of genes increased from 802 to 967 manual annotation actually reduced the number of reactions and metabolites from 1462 to 1137 and 1231 to 1058, respectively.
% The annotation ane experimental evidence from iCac802 \cite{dash2014capturing} to iCac967 \cite{yoo2015quantitative} is greatly expanded.
% However, it is not entirely clear, from the initial publication, the extent to which the expanded iCaC967 model preserves crucial experimental validation.
% The following sections aim to check the later model against previous validation criteria.

% \subsection{Model integration}
% \textbf{Project construction:}
% An initial project of the \textit{C. acetobutylicum} model was created in gsmodutils.
% This is based around the iCac802 base model.
% Included are also knockout strains $\Delta adc, \Delta ack, \Delta ptb$ and $\Delta hbd$ (TODO: describe knockouts).
% 
% In addition, a design listing the constraints changed in model version iCaC967 
% \footnote{We would like to note that such a significant change to a model, in many practical cases, would not be considered a design but, rather, a separate model.
% Separate models within the same project is a feature supported by gsmodutils.}.
% \\
% \textbf{Model comparison: TODO Diffs between existing models - use (and further develop) gsmodutils diff for model stats}.
% 
% \subsection{Evaluation of model validation criteria}
% Several crucial validation criteria were identified from the experimental and theoretical analysis of the models found in the original publications.
% We note that this is not an exhaustive set of criteria.
% This section is designed to serve as a use case for the gsmodutils framework, not to comprehensively review the initial publications.
% \\
% 
% \textbf{TCA cycle}\\
% \textbf{Proton balanced results}\\
% \textbf{Predictions of knockouts}\\
% The inital publication for iCAC802 included validation of four gene knockout mutant strains.
% $\Delta ack$ acetate kinase.\\
% $\Delta ptb$ phosphotransbutrylase.\\
% $\Delta adc$ acetoacetate decarboxylase. \textbf{This reaction is present and the knock out works as expected but not annotated correctly.}\\
% $\Delta hdb$ hydroxybutyryl-CoA dehydrogenase. \\
% 
% \textbf{Flux validation against 13C data}\\
% \textbf{Under acidogenic stress (H inhibited conditions) } \\
% \textbf{Solvent formation Cell recycling in solventogenic phase}\\

\section{Case study usage Clostridium autoethanogenum}
\textit{Clostridium autoethanogenum} is a bacterial species used for the production of commodity chemicals at industrial scale \cite{abrini1994clostridium,norman2018progress,LT}. A new GSMM of \textit{C. autoethanogenum}, `MetaCLAU', has been analysed to improve this bioprocess (Norman et al., unpublished).

In this section we describe how gsmodutils has been utilised to ensure that future versions of MetaCLAU will remain functionally relevant from the perspective of industrial biotechnology.

@@

\textit{C. autoethanogenum} is a strictly anaerobic, acetogenic bacterium which naturally produces ethanol and trace amounts of 2,3-butanediol (2,3BD) from carbon monoxide and water \cite{abrini1994clostridium,schuchmann2014autotrophy,norman2018progress}. Since carbon monoxide is readily available in the form of industrial waste gas, and 2,3BD has a global market value of \$43 billion \cite{kopke201123butanediol}, the optimisation of yields of 2,3BD from carbon monoxide is highly desirable in the context of industry \cite{norman2018progress}.

MetaCLAU was built using Pathway Tools \cite{karp2002pathway} and ScrumPy \cite{poolman2006scrumpy}, and is based on a manually annotated genome sequence of \textit{C. autoethanogenum} \cite{humphreys2015whole}. The resulting model consisted of 758 reactions, 773 metabolites and 518 genes. For full details of the model, see (Norman et al. unpublished)

The model has been integrated with the gsodutils modelling framework as a test-driven project. The following section details specific tests used to evaluate the model at each stage of its continued development.

\subsection{Evaluation of model validation criteria}

\textbf{Energetic consistency:} 
An important limitation of FBA is that optimal solutions may be thermodynamically infeasible if appropriate constraints are not applied \cite{Fell1197rohtua}.
In order to identify these unwanted flux distributions and constrain the model such that they are not in the feasible soluton space, a diagnostic FBA is applied with the following constraints:
1) All transport reactions are constrained to allow no uptake (lower bound = 0), and 2) the ATPase reaction is given a fixed flux of one.
If a (non-zero) solution to this problem exists, it must contain a thermodynamic inconsistency, which can be dealt with by manual inspection of the solution and modification of one or more of the involved reactions \cite{Fell1197rohtua}.

\textbf{Minimisation of flux based tests:}
One conventional approach in FBA is to set an optimisation criteria of minimising flux across enzymatic reactions with a fixed biomass constraint \cite{holzhutter2006thegeneralised}. 
The solution to this FBA problem represents minimal protein investment \cite{holzhutter2006thegeneralised}.
Since execution of flux minimisation in COBRApy requires a model in which reversible reactions are split into two irreversible reactions, a gsmodutils design was created in which all reversible reactions are split in a programmatic manner.
In the case of MetaCLAU, the flux minimal solution includes both ethanol and acetate production, which represents good qualitative agreement with experimental data (Norman et al., unpublished).
Since any changes to this predicted phenotype must be investigated, the flux minimisation analysis has been formulated as a gsmuodutils test.

@@

\textbf{Lethal knockout mutants:} 
The prediction of lethal single-gene KO mutants through FBA of a GSMM is useful in two ways: 
1) the identification of essential genes is an important first step for metabolic engineering strategies, 
and 2) with the advent of high-throughput TraDIS gene-essentiality data sets \cite{langridge2009simultaneous}, GSMMs can be validated by their abiity to predict essential genes.
Furthermore, any change in the set of essential genes (particularly an increase in the number of essentials) represents important information for metabolic engineering.
For these reasons, a test been built into the MetaCLAU project which enables the computation of the set of essential genes and their comparison with TraDIS data sets.

@@

\section{Related work}
\textbf{Memote:} \cite{lieven2017memote} is tool set that has very similar ambitions to gsmodutils.
It features a fully specified set of tests, including custom test cases and has strong version control integration with git.
A core difference between these projects is that gsmodutils has a stronger focus on reducing the redundancy in model storage through the usage of \textit{design deltas}, as described above.
Similarly, a core goal of gsmodutils is to allow easy import and export outside of the framework for compatibility with other modelling suites.
It should be noted that, as memote is written in python and utilises COBRApy, it is fully compatible with gsmodutils.
\\
\textbf{Cameo:} \cite{cardoso2017cameo} is a set of strain design utilities, such as genetic algorithms for finding optimal gene knock-out
strategies, as well as other flux balance analysis utilities. Written in python and part of the COBRApy suite of applications,
gsmodutils integrates easily with cameo.
Indeed, making a set of design changes can easily be integrated into gsmodutils through the python API.
\\
\textbf{Model repositories:} models are frequently shared, at the time of publication through services such as BiGG \cite{king2015bigg} and Biomodels \cite{chelliah2013biomodels}. 
Whilst these repositories encourage the reuse of models and the reproducibility of \textit{in silico} predictions they are not designed to improved collaboration.
The software presented here is designed with the notion that genome scale models are never finished, \textit{per se}, but under continuous development.
The cornerstone of this is the use of test cases, which formalise modelling validation criteria.

\section{Discussion}
In order to facilitate the sharing and dissemination of high quality computational research, good standards and software are required \cite{jimenez2017four}.
Naturally a great deal of effort has gone in to producing high quality systems and synthetic biology standards \cite{hucka2003systems, cox2018synthetic}.
Furthermore, when research projects end it all to is common for important large models to be published and become relics lost within the literature forgotten to all but the most dedicated of individuals.
As GSMMs grow in terms of the metabolism the contain as well as the biological problems they are used to solve, problems with annotation and curation naturally accumulate as a product of human error.
Software that facilitates actively improving how researchers develop and apply models to new phenomena is required.

We have presented a framework with a number of features taken from the software development world specifically designed to improve collaboration and minimise such error.
However, it is important to stress the difference between defined behaviour expected from pre-written test cases and novel predictions made by a model.
Indeed, a core objective of this framework is to ensure that good practices are followed in model development that help scientists to better trust the results discovered by their models.
In an ideal world, we would envision a methodology such as our becoming a pre-requisite for GSMMs to pass peer review.

@@

\section*{Acknowledgements}
We would like to thank the Oxford Brookes Cell Systems Modelling group for helpful discussions regarding this work and Thomas Millat for scientific advice.
This work was supported by the UK Biotechnology and Biological Sciences Research Council (BBSRC) grant BB/L013940/1.

>>>>

\documentclass[a4paper,11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage{fullpage}
\usepackage{color}
\usepackage{authblk}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{hyperref}
%PREPRINT/MANUSCRIPT OPTIONS
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\doublespacing

@@

\title{Gsmodutils: A python based framework for test driven genome scale metabolic model development}
\author[1,2]{James P Gilbert}
\author[1,2]{Nicole Pearcy}
\author[1]{Rupert Norman}
\author[2]{Jamie Twycross}
\author[3]{John King}
\author[4]{Charlie Hodgman}
\author[1]{Nigel Minton}
\affil[1]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\affil[2]{Intelligent Modelling and Analysis group, School of Computer Science, University of Nottingham, Nottingham, NG8 1BB, United Kingdom}
\affil[3]{School of Mathematical Sciences, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\affil[4]{School of Biosciences, University of Nottingham, Sutton Bonington, Loughborough, LE12 5RD, United Kingdom}

@@

\section{Introduction}
Stoichiometric constraints based modelling for biological systems has been a mainstay of systems biology for several decades \cite{fell1986fat, varma1994stoichiometric}.
Given its flexibility, low barrier to entry and requirement only on minimal knowledge regarding the stoichiometry of metabolic networks this structural approach has become an extremely popular method for modelling steady state behaviour of large, biochemical networks \cite{kauffman2003advances}.
Such large scale reconstructions are often referred to as genome scale metabolic models (GSMMs), as the processes is significantly aided through the advent of relatively inexpensive genome sequencing \cite{o2015using, land2015insights}.
Indeed, owing to their ability to model complex aspects of metabolism, GSMMs have been widely adopted as a standard to elucidate and optimise of industrial biotechnology processes \cite{kim2017current}.

@@

Furthermore, as with many areas of bioinformatic study the number available computational tools has become vast.
This covers huge variety of software platforms including the COBRA toolbox in Matlab \cite{schellenberger2011quantitative}, ScrumPy and COBRApy in Python \cite{poolman2006scrumpy, ebrahim2013cobrapy} with additional tools and libraries such as cameo \cite{cardoso2017cameo}, OptFlux in Java \cite{rocha2010optflux} and SurreyFBA \cite{gevorgyan2010surreyfba}.
Whilst most of these tools are Open Source and follow standards, such as SBML \cite{finney2003systems}, there are often few options that make analysis easy to replicate.
The archetype design, build and test cycle of synthetic biology heavily relies on the usage of bioinformatics software and modelling to improve the production of natural products \cite{carbonell2016bioinformatics}.
In order to speed up the use of bioinformatics tools to produce high value platform chemicals, genome scale models are often used to discover methods for process optimisation \textit{in silico}.

For example, many tools such as RetroPath \cite{carbonell2013retropath}, XTMS \cite{carbonell2014xtms} and GEM-Path \cite{campodonico2014generation} suggest thousands of potential heterologous pathways as well as providing predictions for flux distributions.
Many of these tools significantly increase the value of genome scale models, for example by coupling commodity production to an organism's growth \cite{feist2010model}.
These tools all suggest major changes to wild-type strains must be tracked and compared to allow models to remain relevant.
In effect, mechanisms are required to relate modified models to modified test and production strains.

@@

The design phase of industrial synthetic biotechnology, due to historical and regulatory convention, often follows a waterfall model for development of genome scale models.
This is borne out of a number significant constraints in the development of genome scale models.
The waterfall approach for modelling is characterised in Figure \ref{fig:methodology_approaches} where requirements for a model are captured at an early stage and are hard to refine during the production process.

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
Constraints based models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes often being left to a few of critical interest \cite{seemann2014prokka}.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady state phenomena still provides a good understanding of system properties \cite{o2015using}.

@@

\begin{figure}[ht]
\includegraphics[width=\textwidth]{waterfall.eps}
\caption{Conventional waterfall methodology for model development.
Each process is considered to be an isolated aspect of model development.}
\label{fig:methodology_approaches}
\end{figure}

@@

We feel that many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions.
However, there is a major gap in terms of the reliability and re-usability of models due to a lack of standardisation and software tools to aid such processes.
The following sections provide an overview of the gsmodutils software framework.
Gsmodutils aims to provide a basis for test driven, version controlled agile model development.
All software is open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

By automatically integrating COBRApy \cite{ebrahim2013cobrapy} users can easily write convenient test cases following examples given in the user guide.
A standard test case, ensuring that a given model grows on media is given in Figure \ref{fig:test_case}.
When a new model repository is created with the gsmodutils tool, a number of pre-written test cases are automatically added to a file.
However, we stress that the vast majority of individual use cases for a model must be specific to a given biological problem.

The software provides a number of features such as import and export of models in different formats and the generation of test reports through use of the command line. 
The use of flat files enables easy integration with version control software such as git and mercurial.
In addition, projects are easy to export using portable standardised docker images \cite{merkel2014docker}, the idea being to allow users to share models as quickly and easily as possible without concern for custom system configurations.

@@

\subsection{Strain designs}
A core aspect behind the implementation of gsmodutils is the concept of a \textit{design}, this encompasses a simple set of changes to a ``wild-type'' model that are required for analysis.
For example a mutant strain that has undergone several gene deletions is not sufficiently different from the original wild-type to merit having its own genome scale model.
However, it is often the case that such deletions are of scientific or industrial interest and, as such, the strain will be used in future work.
Consequently, such designs are hereditary in nature.
By taking the \textit{model delta} between the constraints applied to an initial model and subsequent modifications, gsmodutils allows users to easily reuse and export models with this design.
As designs inherit from a base model, future curation to a wild type model will automatically be included in the designs.
Similarly, designs are self contained and will not interfere with one another allowing a form of project management.
Figure \ref{fig:strain_hered} shows how this could work in a practical situation.

Designs of this nature can also be programmatic, allowing the implementation of features such as non-standard constraints that can be dynamically loaded.
An example of this is shown in Figure \ref{fig:programmatic_design}.
This example converts an existing model to one based on an MILP, as the reaction names do not need to be specified, should reactions be added or removed the process will still apply.
Alternative examples could include reductions of models through methods such as elementary flux modes or minimum cut sets, which can change dramatically with only small changes to stoichiometry.
Furthermore, integrity of all strain designs is automatically tested as part of the default gsmodutils testing framework.

\begin{figure}[ht]
\lstset{language=Python}   
\begin{lstlisting}
def gsmdesign_example_milp(model, project):
    for reaction in model.reactions:
	reaction.variable_kind = "integer"
    return model

\end{lstlisting}
\label{fig:programmatic_design} 
\caption{An example gsmodutils programmatic design written in python. 
This design converts reactions to integer type, allowing an MILP formation.}
\end{figure}

\begin{figure}
  \includegraphics[width=\textwidth]{inheritence.png}
  \label{fig:strain_hered}
  \caption{Examples of gsmodutils design inheritance.
  Each design stores the delta between the wild-type base model, any parents and the changes to constraints the design contains.
  In the example presented above a heterologous production pathway is combined with a reusable set of knock-outs.
  In practice, ideally, these designs should relate to real constructs and strains evaluated by wet lab biologists.
  }
\end{figure}

\subsection{Development workflow}
In this section we propose a method for the development of genome scale models that integrates gsmodutils with version control systems.
Figure \ref{fig:test_case} highlights the notion of test cases, taken from test driven development.
The outlined workflow is that the user writes a formal test case for some modelling goal, perhaps driven by captured experimental data, that fits a specific form of validation criteria.
We note that, in principle, this test case should be written before changes to a model.

\section{Case study usage Clostridium autoethanogenum}
\textit{Clostridium autoethanogenum} is a bacterial species used for the production of commodity chemicals at industrial scale \cite{abrini1994clostridium,norman2018progress,LT}. A new GSMM of \textit{C. autoethanogenum}, `MetaCLAU', has been analysed to improve this bioprocess (Norman et al., submitted for peer review).
In this section we describe how gsmodutils has been utilised to ensure that future versions of MetaCLAU will remain functionally relevant from the perspective of industrial biotechnology.

@@

\textit{C. autoethanogenum} is a strictly anaerobic, acetogenic bacterium which naturally produces ethanol and trace amounts of 2,3-butanediol (2,3BD) from carbon monoxide and water \cite{abrini1994clostridium,schuchmann2014autotrophy,norman2018progress}. Since carbon monoxide is readily available in the form of industrial waste gas, and 2,3BD has a global market value of \$43 billion \cite{kopke201123butanediol}, the optimisation of yields of 2,3BD from carbon monoxide is highly desirable in the context of industry \cite{norman2018progress}.   
MetaCLAU was built using Pathway Tools \cite{karp2002pathway} and ScrumPy \cite{poolman2006scrumpy}, and is based on a manually annotated genome sequence of \textit{C. autoethanogenum} \cite{humphreys2015whole}. The resulting model consisted of 758 reactions, 773 metabolites and 518 genes. For full details of the model, see (Norman et al. submitted for peer review).
The model has been integrated with the gsodutils modelling framework as a test-driven project. The following section details specific tests used to evaluate the model at each stage of its continued development.

\subsection{Evaluation of model validation criteria}
\textbf{Energetic consistency:} 
An important limitation of FBA is that optimal solutions may be thermodynamically infeasible if appropriate constraints are not applied \cite{Fell1197rohtua}.
In order to identify these unwanted flux distributions and constrain the model such that they are not in the feasible soluton space, a diagnostic FBA is applied with the following constraints:
1) All transport reactions are constrained to allow no uptake (lower bound = 0), and 2) the ATPase reaction is given a fixed flux of one.
If a (non-zero) solution to this problem exists, it must contain a thermodynamic inconsistency, which can be dealt with by manual inspection of the solution and modification of one or more of the involved reactions \cite{Fell1197rohtua}.

\textbf{Minimisation of flux based tests:}
One conventional approach in FBA is to set an optimisation criterion of minimising flux across enzymatic reactions with a fixed biomass constraint \cite{holzhutter2006thegeneralised}. 
The solution to this FBA problem represents minimal protein investment \cite{holzhutter2006thegeneralised}.
Since execution of flux minimisation in COBRApy requires a model in which reversible reactions are split into two irreversible reactions, a gsmodutils design was created in which all reversible reactions are split in a programmatic manner.
In the case of MetaCLAU, the flux minimal solution includes both ethanol and acetate production, which represents good qualitative agreement with experimental data (Norman et al., unpublished).
Since any changes to this predicted phenotype must be investigated, the flux minimisation analysis has been formulated as a gsmuodutils test.

@@

\textbf{Lethal knockout mutants:} 
The prediction of lethal single-gene KO mutants through FBA of a GSMM is useful in two ways: 
1) the identification of essential genes is an important first step for metabolic engineering strategies, 
and 2) with the advent of high-throughput TraDIS gene-essentiality data sets \cite{langridge2009simultaneous}, GSMMs can be validated by their ability to predict essential genes.
Furthermore, any change in the set of essential genes (particularly an increase in the number of essentials) represents important information for metabolic engineering.
For these reasons, a test has been built into the MetaCLAU project which enables the computation of the set of essential genes and their comparison with TraDIS data sets.

@@

\section{Related work}
\textbf{Memote:} \cite{lieven2017memote} is tool set that has very similar ambitions to gsmodutils.
It features a fully specified set of tests, including custom test cases and has strong version control integration with git.
A core difference between these projects is that gsmodutils has a stronger focus on reducing the redundancy in model storage through the use of \textit{design deltas}, as described above.
Similarly, a core goal of gsmodutils is to allow easy import and export outside of the framework for compatibility with other modelling suites.
It should be noted that, as memote is written in python and utilises COBRApy, it is fully compatible with gsmodutils.
\\
\textbf{Cameo:} \cite{cardoso2017cameo} is a set of strain design utilities, such as genetic algorithms for finding optimal gene knock-out
strategies, as well as other flux balance analysis utilities. Written in python and part of the COBRApy suite of applications,
gsmodutils integrates easily with cameo.
Indeed, making a set of design changes can easily be integrated into gsmodutils through the python API.
\\
\textbf{Model repositories:} models are frequently shared, at the time of publication through services such as BiGG \cite{king2015bigg} and Biomodels \cite{chelliah2013biomodels}. 
Whilst these repositories encourage the reuse of models and the reproducibility of \textit{in silico} predictions they are not designed to improved collaboration.
The software presented here is designed with the notion that genome scale models are never finished, \textit{per se}, but under continuous development.
The cornerstone of this is the use of test cases, which formalise modelling validation criteria.

\section{Discussion}
In order to facilitate the sharing and dissemination of high quality computational research, good standards and software are required \cite{jimenez2017four}.
Naturally a great deal of effort has gone in to producing high quality systems and synthetic biology standards \cite{hucka2003systems, cox2018synthetic}.
Furthermore, when research projects end it all to is common for important large models to be published and become relics lost within the literature, forgotten to all but the most dedicated of individuals.
As GSMMs grow in terms of the metabolism the contain as well as the biological problems they are used to solve, problems with annotation and curation naturally accumulate as a product of human error.
Software that facilitates actively improving how researchers develop and apply models to new phenomena is required.

We have presented a framework with a number of features taken from the software development world specifically designed to improve collaboration and minimise such error.
However, it is important to stress the difference between defined behaviour expected from pre-written test cases and novel predictions made by a model.
Indeed, a core objective of this framework is to ensure that good practices are followed in model development that help scientists to better trust the results discovered by their models.
In an ideal world, we would envision a methodology such as ours becoming a pre-requisite for GSMMs to pass peer review.

@@

\section*{Acknowledgements}
We would like to thank the Oxford Brookes Cell Systems Modelling group for helpful discussions regarding this work and Thomas Millat for scientific advice.
This work was supported by the UK Biotechnology and Biological Sciences Research Council (BBSRC) grants BB/L013940/1,  BB/K00283X/1 and BB/L502030/1.
